Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-16 doc/mc-data.tex
Filled out experimental results

Ran all benchmarks for which generation + checking could complete in 1000 seconds.  Total = 82.  All completed in less than 1000 seconds with structured.
Compare both runtimes and total number of clauses for structured vs. monolithic:
\begin{itemize}
\item Structured complete 82.
\item Monolithic completed  69, timed out on 13.
\item Of the 69 completed, 23 were faster with monolithic, 36 were slower (min ratio = 0.015, max ratio = 9.66, hmean = 0.40)
\item Of the 69 completed, 59 had fewer total clauses with monolithic, 10 had more (min ratio = 0.07, max ratio = 1.09, hmean = 0.49)
\end{itemize}

>>>>

\input{figure-clauses-monolithic}

Ran all benchmarks for which generation + checking could complete in 1000 seconds.  Total = 82.  All completed in less than 1000 seconds with structured.
Compare both runtimes and total number of clauses for structured vs. monolithic:
\begin{itemize}
\item Structured complete 83.
\item Monolithic completed  70,timed out on 13.
\item Of the 70 completed, 23 were faster with monolithic, 37 were slower (min ratio = 0.015, max ratio = 9.66, hmean = 0.40)
\item Of the 70 completed, 60 had fewer total clauses with monolithic, 10 had more (min ratio = 0.07, max ratio = 1.09, hmean = 0.49)
\end{itemize}

### 2023-03-16 doc/mc-paper.tex
Filled out experimental results

With a runtime limit of 4000 seconds (the model counting competition
has a limit of 3600 seconds), \dfour{} completed for 124 of the
benchmark problems.  Our proof generator was able to convert all of
these into POGs, with their declarations ranging from 304 to
2,761,457,765 (median 848,784) defining clauses.
%(The
%maximum count overflowed the overflowed the 32-bit signed integers we
%used to represent clause identifiers in both the proof generator and
%checker.)
We then ran our proof generator with a time limit of 10,000 seconds.
It was able to generate full proofs for 108 of the problems and
one-sided proofs for an additional 9 of them, leaving just 7 with no
verification.  The checker successfully verified all of the generated
proofs.  The longest runtime for the combination of proof generator
and checker for a full proof was 13,145 seconds.
It is worth noting that, to date, we have not found
any errors in the d-DNNF graph generated by \dfour{}.

@@

\section{Optimizations for Forward Implication Proofs}
\label{app:optimizations}

@@

A single recursive step of $\validate$ can encounter product nodes
having many---tens or even hundreds---of literals as children.  The earlier formulation of $\validate$
considers each literal $\lit \in L$ separately,
calling a SAT solver
on the formula  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ when $\lit$ cannot be validated with BCP\@.

Literal grouping allows all literals to be validated with a single call to the SAT solver.
It collects those literals
$\lit_1, \lit_2, \ldots, \lit_m$ that cannot be validated by BCP and defines a
product node $\nodev$ having these literals as children.  The goal
then becomes to prove that any total assignment must yield 1 for extension
variable $v$.  A single call to the solver can generate this proof by invoking it on the formula
  $\simplify{\psi}{\passign} \cup \simplify{\theta_{\nodev}}{\{ \obar{v} \}}$ to show that is unsatisfiable.
  The proof steps can be mapped back into clause addition steps in the CRAT file, using as hints the
  input clauses and the defining clauses for $\nodev$.

@@

@@

@@

@@

@@

\input{figure-mice-crat}

>>>>

With a runtime limit of 4000 seconds (the model counting competition
has a limit of 3600 seconds), \dfour{} completed for 124 of the
benchmark problems.  Our proof generator was able to convert all of
these into POGs, with their declarations ranging from 304 to
2,761,457,765 (median 848,784) defining clauses.
%(The
%maximum count overflowed the overflowed the 32-bit signed integers we
%used to represent clause identifiers in both the proof generator and
%checker.)
We then ran our proof generator with a time limit of 10,000 seconds.
It was able to generate full proofs for 108 of the problems and
one-sided proofs for an additional 9 of them, leaving just 7 with no
verification.  The checker successfully verified all of the generated
proofs.  The longest runtime for the combination of proof generator
and checker for a full proof was 13,145 seconds.
It is worth noting that, to date, we have not found
any errors in the d-DNNF graphs generated by \dfour{}.

@@

\section{Optimizations for Forward Implication Proofs}

The two optimizations we have implemented exploit the power of our
general resolution framework to define new structures within the
proof. In particular, they create new product nodes that are not part
of the POG representation.

\label{app:optimizations}

@@

A single recursive step of $\validate$ can encounter product nodes
having many literals as children.  The naive formulation of $\validate$
considers each literal $\lit \in L$ separately.
Literal grouping allows all literals to be validated with a single call to a SAT solver.
It collects those literals
$\lit_1, \lit_2, \ldots, \lit_m$ that cannot be validated by BCP and defines a
product node $\nodev$ having these literals as children.  The goal
then becomes to prove that any total assignment must yield 1 for extension
variable $v$.  A single call to the solver can generate this proof by invoking it on the formula
  $\simplify{\psi}{\passign} \cup \simplify{\theta_{\nodev}}{\{ \obar{v} \}}$ to show that is unsatisfiable.
  The proof steps can be mapped back into clause addition steps in the CRAT file, using as hints the
  input clauses and the defining clauses for $\nodev$.

@@

The primary purpose of our experiments is to assess the ability of our
implementation of the toolchain illustrated in Figure~\ref{fig:chain}
to check the results from a possible knowledge compiler competition,
along the lines of the annual competitions in SAT solvers, model
counters, and other automated reasoning tools.  As secondary goals, we
seek insights into how possible algorithmic choices affect the
performance of the toolchain and how our toolchain compares to the
MICE framework for verifying model counters~\cite{fichte:sat:2022}.

\subsection{Experimental Setup}

As a test machine, we used a 2021~MacBook~Pro laptop having a 3.2~Ghz
Apple M1 processor and with 64~GB of physical memory.  We also used an
external 3~TB solid-state disk drive with an advertised read/write
speed of one gigabit per second.  Despite being a laptop, this
configuration is somewhat more powerful than the nodes typically
found in server clusters.  The fast access to storage is especially
important due to the very large files generated and processed by the
tools.  Our tools generated individual files with over 160 gigabytes
of data.

For benchmarks, we downloaded 100 files each from the 2022 standard and weighted model counting competitions:
\begin{center}
       \url{https://mccompetition.org/2022/mc_description.html}
\end{center}
We found that 20 of these were duplicates across the two tracks,
yielding a total of 180 unique benchmark problems, ranging in size
from 250 to 2,753,207 clauses.

Our default configuration for the proof generator used the structural
method for the forward implication proof, with the two optimizations
of literal grouping and lemmas.
Running with a time limit of 4000 seconds\footnote{All measured times
in this document are actual elapsed times, not CPU times.}, \dfour{}
was able to generated d-DNNF representations for 124 of these, and it timed out on the
rest.  Our proof generator was able to convert all of the generated
d-DNNF graphs into POGs and determine how many defining clauses they
would generate.  A POG operation with $k$ arguments requires $k+1$
defining clauses, and so the total number of defining clauses is equal
to the sum of the number of nonterminal nodes plus the sum of number
of children for each of these nodes.  The number of defining clauses
ranged from 304 to 2,761,457,765 with a median of 848,784.

\subsection{Performance of the Proof Generator and Checker}

@@

For each of the 124 d-DNNF graphs generated by \dfour{}, we attempted
to use our proof generator and unverified checker, limiting proof
generation to 10,000 seconds.  We also attempted to generate and check
one-sided proofs for 124 graphs, again with a time limit of 10,000
seconds for proof generation.

Figure~\ref{fig:d4:crat} summarizes our results in terms of the time
required by \dfour{} (X axis) versus the sum of the times for the
proof generator and checker (Y axis).
We were able to complete a full validation for 108 of the 124
benchmarks, with times ranging from fractions of a second to 13,144.
seconds, with a median of 71.6 seconds.

Relative to the runtime for \dfour{}, two problems ran faster
with the generator and checker.  These were for problems having small numbers of
models (relative to the number of variables), and so most of the time
spent by both \dfour{} and our proof generator was in running a CDCL
solver to search the very sparse solution space.  Our solver,
\cadical{} generally outperforms the miniSAT-based CDCL solver used by
\dfour{}.  At the other extreme, one problem that required only 19
seconds for \dfour{} required 8657 seconds to generate a proof and 45
seconds to check it.  It is unclear why the proof generation required
so much time.  Overall the ratios between the combined generation plus
checking times versus the time for \dfour{} had a harmonic mean of
5.5.

Of the 16 benchmarks that could not be fully validated, one had so
many defining clauses that it overflowed the 32-bit signed integers
our programs used for clause identifiers.  Another exited due to the
virtual memory limit imposed by the operating system, and the
other 14 timed out.  Of the 15 that did not overflow the clause
counter, we were able to complete a one-sided verification for 9, but
the other 6 timed out.

@@

Figure~\ref{fig:defining:total} compares the number of defining
clauses (X axis) with the total number of clauses (Y axis) for the 108
problems that were fully verified.  These totals include the defining
clauses for the POG, the additional defining clauses introduced to
support literal grouping and lemmas, and the clauses added by RUP
steps in the forward implication proof.  These totals ranged from 554
to 770,382,773, with a median of 1,719,245.  The ratios of the total
clauses versus definining clauses ranged from 1.54 to 6073, with a
harmonic mean of 3.13.  The high numbers were for problems having very
few models, and so the proof clauses must encode large proofs of
unsatisfiability.

\subsection{Monolithic Forward Implication Proofs}

@@

\input{figure-clauses-monolithic}

To assess the relative merits of the two approaches to forward
implication proof generation, we ran the proof generator in monolithic
mode on the 92 problems for which the combination of structural proof generation
and checking required at most 3000 seconds and then kept only those
measurements for which at least one of the two approaches had combined
times (generation plus checking) of at most 1000 seconds.  There were
a total of 83 such problems.  Figures~\ref{fig:monolithic:time} and
\ref{fig:monolithic:clauses} show comparisons between the two
approaches in terms of time and total clauses.

Examining Figure~\ref{fig:monolithic:time}, we can see that the
monolithic approach ran faster than the structural approach for 23 of
the problems, including a case where the generator and checker
required only one second, versus 71 seconds for the structural
approach.  On the other hand, the monolithic approach timed out for 13
of the cases and was slower for the other 37.  In general, we can see
the monolithic approach faring worse on larger problems.

Figure~\ref{fig:monolithic:clauses} shows the total numbers of clauses
for the two approaches for the 70 cases where both approaches completed within 1000 seconds.
As can be seen, the monolithic approach consistently produces
shorter proofs, perhaps due to the benefits of the proof trimming by
\dtrim{}.

These experiments suggest that a hybrid of the two approaches might
yield the best results in terms of consistency, runtime, and proof
size.  It would begin the recursive descent of $\validate$ according
to a structural approach, but monolithically generate forward
implication proofs once it encounters sufficiently small subgraphs.

\subsection{Impact of Optimizations}

@@

To assess the impact of the two optimizations: literal grouping and
lemmas, we ran the proof generator on the 80 problems for which the
combination of the two optimizations yielded proofs totaling at most
100~million clauses.  Compared to the problems used to assess monolithic
proof generation, this set included one that had a short proof but
long runtime and excluded four that had short runtimes but long
proofs.  We then ran the proof generator on all three combinations of
the optimizations being partially or totally disabled.
Figure~\ref{fig:optimized:lessoptimized} shows how disabling these optimizations affected the
total number of clauses in the proofs, where the X axis indicates the
total number with both optimizations enabled, while the Y axis
indicates the number with one (left) or both (right) optimizations
enabled.

As the figures illustrated, each optimization on its own can shorten
the proofs, sometimes substantially, and the two combine to have even
greater effect.  Of the 80 benchmark problems evaluated, and comparing
with both optimizations disabled, 47 had the number of proof clauses
reduced by a factor of 2 by using lemmas, 14 by using literal merging,
and 60 by enabling both optimizations.  In the extreme cases, a lack
of lemmas caused one proof to grow by a factor of 52.5, while a lack
of literal grouping causes another proof to grow by a factor of 39.6.

The time performance of these optimizations is less dramatic, but
still significant.  Lemmas sometimes made proof generation slower, but
in one case ran $38\times$ faster.  Similarly, literal grouping could
take longer, but in one case ran $3.9\times$ faster.

Overall, it is clear that these two optimizations are worthwhile, and
even critical to success for some benchmarks.

\subsection{Comparison with MICE Framework}

To compare the performance of our toolchain versus that of
MICE~\cite{fichte:sat:2022}, we consider how the combination of proof
generation, proof checking, and ring evaluation in the CRAT toolchain compare to the time to
run their proof generator \progname{nnf2trace} and checker
\progname{sharptrace}.  We add the time for ring evaluation on our
side, since their checker provides a count, although only an
unweighted one.  We ran their tools on the 92 benchmark problems for
which our toolchain requires at most 3000 seconds and retained
the 84 data points for which at least one of the toolchains completes in under 1000 seconds.

We encountered a significant number of errors running their tools.
These seem mostly to be easily fixable, and we are currently
supporting their debugging efforts.  At present, we can summarize the results as:
\begin{itemize}
\item 84 total data points
\item 67 were completed by both programs in under 1000 seconds
\item Of those 21 were faster with MICE, 46 with CRAT
\item 7 completed with CRAT under 1000 seconds but exceeded that threshold for MICE
\item 1 completed with MICE under 1000 seconds but exceeded that threshold for CRAT
\item 9 completed with CRAT but generated errors for MICE.
\end{itemize}

\input{figure-mice-crat} The detailed results are shown in
Figure~\ref{fig:mice}.  As can be seen, the comparative runtimes are
highly variable, reflecting the fact that the two toolchains differ
fundamentally in their objectives and their approaches.

One shortcoming of \progname{nnf2trace} can be seen in
the near-vertical series of points in the upper-lefthand corner.
These correlate with a similar set of points along the left in
Figure~\ref{fig:optimized:lessoptimized}.  Like the naive
implementation of $\validate$, their program recursively traverses the
graph representation of a formula, effectively expanding it into a
tree.  They lack an optimization analogous to lemmas.

We have not tried the MICE toolchain on larger benchmarks, but we
suspect it would encounter significant scaling limitations.
